Supporting Information : The importance of a complete statistical description of dependence between variables for ecology and related fields

### 2019-04-27 Paper.Rmd
results and description added for Holder's evol mechanism

<!--Holder stuff-->
Our simulations of character evolution suggest the hypothesis that evolutionary
changes in bird and mammal BMR and body size may exhibit right-tail dependence. This is
a hypothesis only, because the tail dependence of Fig. \ref{fig_biv_multi_raw_cop}F,G
could have come about in another, unknown way: our simulations show that tail dependence 
in evolutionary changes is sufficient, but may not be necessary, to produce tail
dependence in characters of extant species. Even if the hypothesized mechanism
is correct, our results simply replace one question, i.e., why do we see right-tail
dependence between BMR and body mass for birds and mammals, with another question,
i.e., why might we see right-tail dependence in evolutionary changes of these traits.
See Appendix \ref{evol_mech} for additional thoughts on the potential importance of 
copulas for studies of character evolution when phylogeny branch lengths are not available.

>>>>

<!--Holder stuff-->
Our simulations of character evolution suggest the hypothesis that evolutionary
changes in bird and mammal BMR and body size may exhibit right-tail dependence. This is
a hypothesis only, because the tail dependence of Fig. \ref{fig_biv_multi_raw_cop}F,G
could have come about in another, unknown way: our simulations show that tail dependence 
in evolutionary changes is sufficient, but may not be necessary, to produce tail
dependence in characters of extant species. Even if the hypothesized mechanism
is correct, our results simply replace one question, i.e., why do we see right-tail
dependence between BMR and body mass for birds and mammals, with another question,
i.e., why might we see right-tail dependence in evolutionary changes of these traits.
See Appendix \ref{SM-evol_mech} for additional thoughts on the potential importance of 
copulas for studies of character evolution when phylogeny branch lengths are not available.

### 2019-04-27 SupportingInformation.Rmd
results and description added for Holder's evol mechanism

Once character evolution was simulated 100 times for each dependence structure, 
we had 817 bivariate characters for each of 500 simulations. 
<!--***DAN: Shya to insert as the rest of this paragraph a description of what 
you did to each of these datasets to assess symmetry or asymmetry of tail dependence-->

<!--***DAN: Shya, please insert a new para here with a description of results, including a new figure with 5 panels showing the 5 histograms you showed me.-->

@@

>>>>

Once character evolution was simulated 100 times for each dependence structure, 
we had 817 bivariate characters for each of 500 simulations. We generated 500 bivariate copulas 
each made of 817 points. To assess the symmetry or asymmetry of tail dependence, we computed 
our three non-parametric statistics ($\cor$, $\Ps$, $\Dsq$) on both extreme tips (lower, henceforth subscripted with `l` 
and upper, henceforth subscripted with `u`) of each copula.
To measure lower tail and upper tail statistics we used the bounds as $(l_b=0,u_b=0.2)$ and 
$(l_b=0.8,u_b=1)$, respectively. Therefore, positive (or negative) difference of tail dependence statistics
($\cor_l-\cor_u$, $\Ps_l - \Ps_u$ and $\Dsq_u - \Dsq_l$) would indicate asymmetry in dependence structure with
stronger association in lower tail (or upper tail) of the bivariate copula.

Five separate histograms were made with those 500 $\cor_l-\cor_u$ statistics, where each 100 
sets were representative of 
five different groups of simulations. These five groups of simulations used random draws from
1) extreme left-tail, 2) moderate left-tail,
3) symmetric tail (bivariate `N` copula), 4) moderate right-tail,
and 5) extreme right-tail dependence copula, respectively, as two traits evolved through each branch.
Five histograms each consisting 100 $\Ps_l - \Ps_u$ statistics and another five histograms each consisting
100 $\Dsq_u - \Dsq_l$ statistics were produced in a similar way. Fig. \ref{fig_cop_evo} represents 
three statistics for each five sets of dependence structure. If most of the statistics showed major
occurrence on either side of zero line (vertical dashed line in Fig. \ref{fig_cop_evo}), then
asymmetry in copula structure was confirmed. Fig. \ref{fig_cop_evo}(A-C, D-F) showing 
stronger lower tail dependence with three statistics majorly distributed 
on positive side were consistent with left tail-dependent copula as mentioned (1) and (2), respectively. 
Similarly, Fig. \ref{fig_cop_evo}(J-L, M-O) showing 
stronger upper tail dependence with three statistics majorly distributed 
on negative side were consistent with right tail-dependent copula as mentioned (4) and (5), respectively.
Evenly distributed tail dependence-difference statistics (see Fig. \ref{fig_cop_evo}G-I) 
around zero line was indicative of symmetric dependence as emerged from bivariate normal copula.

@@

<!--Following chunk is to examine the hypothesis for third probable mechanism : 
asymmetric tail dependence in evolutionary changes in bivariate characters can give
rise to similarly asymmetric tail dependence in character distributions across phylogeny tips-->

```{r mechanism_evo_cop, echo=F, results="hide", cache=T, cache.extra=list(seed,mtime("explore_evo_cop.R"),mtime("CopulaFunctions_flexible.R"))}
set.seed(seed)

source("explore_evo_cop.R")
```

<!--adding plots showing results to confirm hypothesis behind third mechanism-->
\begin{figure}[!h]
\textbf{ \hspace{6.5 cm} (A) \hspace{2 cm} (B) \hspace{2 cm} (C)} \\
\vspace{-0.5 cm}
\begin{center}
\includegraphics[width=9 cm]{./Results/BMR_results/evo_cop/ExtremeL_hist_taildep.pdf}\\
\end{center}
\textbf{ \hspace{6.5 cm} (D) \hspace{2 cm} (E) \hspace{2 cm} (F)} \\
\vspace{-0.5 cm}
\begin{center}
\includegraphics[width=9 cm]{./Results/BMR_results/evo_cop/SomewhatL_hist_taildep.pdf}\\
\end{center}
\textbf{ \hspace{6.5 cm} (G) \hspace{2 cm} (H) \hspace{2 cm} (I)} \\
\vspace{-0.7 cm}
\begin{center}
\includegraphics[width=9 cm]{./Results/BMR_results/evo_cop/BivNormal_hist_taildep.pdf}\\
\end{center}
\textbf{ \hspace{6.5 cm} (J) \hspace{2 cm} (K) \hspace{2 cm} (L)} \\
\vspace{-0.5 cm}
\begin{center}
\includegraphics[width=9 cm]{./Results/BMR_results/evo_cop/SomewhatR_hist_taildep.pdf}\\
\end{center}
\textbf{ \hspace{6.5 cm} (M) \hspace{2 cm} (N) \hspace{2 cm} (O)} \\
\vspace{-0.5 cm}
\begin{center}
\includegraphics[width=9 cm]{./Results/BMR_results/evo_cop/ExtremeR_hist_taildep.pdf}\\
\caption[Non-parametric statistics to check probable mechanism behind asymmetric tail dependence in character distributions across phylogeny tips]{Three measures of asymmetric tail dependence using non-parametric statistics ($\cor,\Ps$ and $\Dsq$)
on lower ($l_b=0,u_b=0.2$) and upper ($l_b=0.8,u_b=1$) tips confirms our hypothesis that copula structure in evolutionary changes in bivariate characters can give rise to similar tail dependent copula in character distributions across phylogeny tips (for details see Appendix \ref{evol_mech}). Stronger lower tail dependence (distributed more on positive values) was found in bivariate traits when (A-C) extreme left tail and (D-F) moderately left tail dependent copula was used in evolutionary shifts through each branch. Similarly, non-parametric statistics showed symmetric tail dependence for (G-I) normal copula and stronger upper tail dependence for (J-L) moderately right tail and (M-O) extreme right tail dependent copula, respectively.\label{fig_cop_evo}}
\end{center}
\end{figure}
